NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2019-02-11 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPSvsPC.eps}
  \caption{\label{HGorderparametersPSvsPC}
    Changes of POPS headgroup order parameters with increasing amount of POPC in POPC:POPS mixtures at 298~K.
    Experimental values with the signs are measured for pure POPS system in this work.
    The signs are assumed to be the same for the mixture and the values are from Ref. \citenum{roux90}.
%    Since the experimental data of POPS in pure and diluted mixture come from
%    different experimental sets ($^{13}$C NMR in this work and $^{2}$H NMR in Ref. \citenum{roux90}),
%    the experimental change of the order parameter is less accurate than in typical measurements
%    where same technique is used in all conditions, see discussion about qualitative and quantitative 
%    accuracy in Ref. \citenum{ollila16}.
    The y-axis for the $\alpha$-carbon results of POPS (bottom) is transferred
    with the same value for both order parameters such that the lower order
    parameter value from pure POPS is at zero to correctly illustrate the significant forking.
  }
\end{figure}

Membranes containing PS lipids are always accompanied with counterions that
modulate electrostatic interactions between lipids and other biomolecules. 
Counterions are also suggested to screen the repulsion between charged lipid headgroups 
in MD simulations and thus to reduce the area per lipid of PS bilayers to be smaller than in PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06}. 
Counterion density profiles along membrane normal 
indicate significant differences between force fields in both binding affinity
and distribution of ions in the interface (Fig. \ref{NAdensPOPS}).
The experimental area per lipid (62.7~\AA$^2$) \cite{pan14} 
is reproduced only in Gromos-CKP and in the MacRog simulation
with potassium counterions, while other models give significantly lower areas (Fig. \ref{NAdensPOPS}).
The counterion binding and the concomitant electrostatic screening of the headgroup
repulsion does not fully explain the low area per molecule values,
because the MacRog simulation, which has the strongest sodium binding
(the lowest concentrations in bulk water), gives the same area per molecule
as the CHARMM36ua simulation, which has significantly weaker counterion binding
affinity. On the other hand, changing counterions from sodium to potassium, having weaker binding
affinity, increases the area per molecule from 53~\AA$^2$ to 63~\AA$^2$ in MacRog simulations.
In conclusion, the results are in line with the previous study suggesting that the
low areas per molecule in PS lipid bilayers originate from the combination
of both counterion binding and hydrogen bonding network between lipid
headgroups \cite{petrache04}.

\begin{figure*}[ht]
  \centering
  \includegraphics[width=18cm]{../Figs/CHANGESwithCaClPS.eps}
  \caption{\label{changesWITHCaClPS}
    Changes of POPC (left) and POPS (right) headgroup order parameters from POPC:POPS (5:1) mixture
    as a function CaCl$_2$ concentration from experiments \cite{roux90} and different simulations
    at 298K (except the data for Berger model is from simulation of POPC:POPS (4:1) mixture at 310K \cite{ollila07a,melcrova16}). 
    The order parameter values from systems without calcium are set as the zero point of y-axis,
    except for the $\alpha$-carbon order parameter of POPS (bottom, right) for which the both order parameters are shifted
    such that the lower order parameter is zero without additional ions to correctly illustrate
    the forking with different concentrations of calcium.
    Potassium counterions are used in MacRog simulations and sodium counterions in Lipid14/17 simulations.
    In CHARMM36 and Berger simulation with added calcium, the charge is neutralized with calcium and monovalent counterions are not present.
  }
%  \todo{Upcoming simulations with original CHARMM36 have been mentioned in the blog:
%    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1520090718976\#c5569269391707740056.
%  these are not necessary, but could be added here if delivered.} \\
\end{figure*}

Binding of cations to zwitterionic PC lipid bilayers has been previously
evaluated against experiments using the changes of headgroup order parameters
as a function of salt concentration \cite{catte16}.
Studying binding of cations to negatively charged lipid bilayers is less straighforward,
because the cationic counterions are always present and the ion-free reference state does thus not exist.
In addition, the analysis is complicated by the artificial aggregation of counterions
in solution observed in some simulations (section \ref{mixtureTOadditionalCIs} in the supplementary information).
Therefore, we evaluate here the amount of bound charge not by adding salt
(although also this is discussed in the section \ref{mixtureTOadditionalCIs} in the supplementary information),
but by studying the changes of the headgroup order parameters with increasing amount of
negatively charged lipids (and thus increasing amount of cationic counterions) in the bilayer.
According to the molecular electrometer concept, the headgroup order parameters of POPC
increase when negatively charged POPS lipids are incorporated in lipid bilayer
(section \ref{HGorderparametersPCvsPEPSPGchol})~\cite{seelig87,scherer87}.
This is reproduced in the MacRog simulations with potassium counterions (Fig. \ref{HGorderparametersPCvsPS}),
which have the weakest binding affinity to POPS lipid bilayers (Fig.~\ref{NAdensPOPS}).
The CHARMM36 and Berger simulations predict no change, or a decrease,
in the POPC headgroup order parameters as a function of increased amount of POPS (Fig. \ref{HGorderparametersPCvsPS}).
This can be explained by the stronger counterion binding affinity, which cancels
the effect of negatively charged headgroups and prevents the experimentally observed
increase of headgroup order parameters with increasing amount of PS lipids.
Therefore, we suggest that the relatively weak binding of potassium
in the MacRog simulations (Fig. \ref{NAdensPOPS}) predicts the most
realistic surface charge density in membranes containing PS lipids,
while the other tested simulation models overestimate the counterion
binding affinity. The results are in line with the changes of headgroup order
parameters as a function of added counterions analyzed in section \ref{mixtureTOadditionalCIs}
in the supplementary information.

The reduced forking of the POPS $\alpha$-carbon (Fig.~\ref{HGorderparametersPSvsPC})
together with other experimental results suggest less rigid structure of PS headgroups when diluted with
POPC~\cite{browning80,buldt81,roux90,roux91,scherer87}.
None of the tested models reproduce the changes of POPS headgroup order
parameters with increasing amount of POPC in POPC:POPS mixtures (Fig. \ref{HGorderparametersPSvsPC}).
Therefore, we conclude that more accurate force fields are necessary
to correctly describe the PC-PS headgroup interactions in MD simulations.

>>>>

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPSvsPC.eps}
  \caption{\label{HGorderparametersPSvsPC}
    Modulation of POPS headgroup order parameters with increasing amount of POPC in POPC:POPS mixtures at 298~K.
    Experimental values with the signs are measured for pure POPS system in this work.
    The signs are assumed to be the same for the mixture and the values are from Ref. \citenum{roux90}.
%    Since the experimental data of POPS in pure and diluted mixture come from
%    different experimental sets ($^{13}$C NMR in this work and $^{2}$H NMR in Ref. \citenum{roux90}),
%    the experimental change of the order parameter is less accurate than in typical measurements
%    where same technique is used in all conditions, see discussion about qualitative and quantitative 
%    accuracy in Ref. \citenum{ollila16}.
    The y-axis for the $\alpha$-carbon results of POPS (bottom) is transferred
    with the same value for both order parameters such that the lower order
    parameter value from pure POPS is at zero to correctly illustrate the significant forking.
  }
\end{figure}

Membranes containing PS lipids are always accompanied with counterions that
modulate electrostatic interactions between lipids and other biomolecules. MD simulations have suggested
that counterions reduce the area per lipid of PS bilayers to be smaller than in PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06} by screening the repulsion between charged lipid headgroups. 
Counterion density profiles along membrane normal 
show significant differences between force fields in both binding affinity
and distribution of ions in the interface (Fig. \ref{NAdensPOPS}).
The experimental area per lipid (62.7~\AA$^2$) \cite{pan14} 
is reproduced only in Gromos-CKP and in the MacRog simulation
with potassium counterions, while other models give considerably smaller areas (Fig. \ref{NAdensPOPS}).
However, the counterion binding and the concomitant electrostatic screening of the headgroup
repulsion does not fully explain the low area per molecule values.
since the MacRog simulation, which has the strongest sodium binding
(the lowest concentrations in bulk water), gives the same area per molecule
as the CHARMM36ua simulation, which has significantly weaker counterion binding
affinity. On the other hand, MacRog simulations with potassium produce a larger area per molecule (63~\AA$^2$) than with natrium (53~\AA$^2$) in line with the weaker potassium binding affinity (Fig.~\ref{NAdensPOPS}).
In conclusion, the results are in line with the previous study suggesting that the
low areas per molecule in PS lipid bilayers originate from the combination
of both counterion binding and hydrogen bonding network between lipid
headgroups \cite{petrache04}. \todo{can we really say this based on our data? We don't discuss hydrogen bonding}

\begin{figure*}[ht]
  \centering
  \includegraphics[width=18cm]{../Figs/CHANGESwithCaClPS.eps}
  \caption{\label{changesWITHCaClPS}
   Variation of POPC (left) and POPS (right) headgroup order parameters from POPC:POPS (5:1) mixture
    as a function CaCl$_2$ concentration from experiments \cite{roux90} and different simulations
    at 298K (except the data for Berger model is from simulation of POPC:POPS (4:1) mixture at 310K \cite{ollila07a,melcrova16}). 
    The order parameter values from systems without calcium are set as the zero point of y-axis,
    except for the $\alpha$-carbon order parameter of POPS (bottom, right) for which both order parameters are shifted
    such that the lower order parameter is zero without additional ions. This is to correctly illustrate
    the forking with different concentrations of calcium.
    Potassium counterions are used in MacRog simulations and sodium counterions in Lipid14/17 simulations.
    In CHARMM36 and Berger simulation with added calcium, the charge is neutralized with calcium and monovalent counterions are not present.
  }
%  \todo{Upcoming simulations with original CHARMM36 have been mentioned in the blog:
%    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1520090718976\#c5569269391707740056.
%  these are not necessary, but could be added here if delivered.} \\
\end{figure*}

The experimentally observed modulation of headgroup order parameters
by increasing salt concentration (the molecular electrometer concept) has been previously used to evaluate the cation binding to zwitterionic PC bilayers in simulations \cite{catte16}.
Studying binding of cations to negatively charged lipid bilayers is less straighforward due to the presence of cationic counterions and the lack of a ion-free reference state. The analysis is further complicated by the artificial aggregation of counterions
observed in some simulations (section \ref{mixtureTOadditionalCIs} in the SI)\todo{Why isn't this a problem with PC}.
Therefore, we evaluate here the amount of bound charge not by adding salt
(although also this is discussed in the SI section \ref{mixtureTOadditionalCIs}),
but by studying the changes of the headgroup order parameters with increasing amount of
negatively charged lipids (and thus increasing amount of cationic counterions) in the bilayer.

According to the molecular electrometer concept, the headgroup order parameters of POPC
increase when negatively charged POPS lipids are incorporated in lipid bilayer
(section \ref{HGorderparametersPCvsPEPSPGchol})~\cite{seelig87,scherer87}.
This is reproduced in the MacRog simulations with potassium counterions (Fig. \ref{HGorderparametersPCvsPS}),
which have the weakest binding affinity to POPS lipid bilayers (Fig.~\ref{NAdensPOPS}).
The CHARMM36 and Berger simulations either exhibit no change or show a decrease
in the POPC headgroup order parameters as the amount of POPS increases (Fig. \ref{HGorderparametersPCvsPS}).
This can be explained by the stronger counterion binding affinity, which cancels
the effect of negatively charged headgroups and prevents the experimentally observed
increase of headgroup order parameters with growing amount of PS lipids.
Therefore, we suggest that the relatively weak binding of potassium
in the MacRog simulations (Fig. \ref{NAdensPOPS}) produces the most
realistic surface charge density in membranes containing PS lipids,
while the other tested simulation models overestimate the counterion
binding affinity. The results are in line with the behaviour of headgroup order
parameters as a function of added counterions analyzed in section \ref{mixtureTOadditionalCIs}
in the SI.

The reduced forking of the POPS $\alpha$-carbon (Fig.~\ref{HGorderparametersPSvsPC})
together with other experimental results suggest less rigid structure of PS headgroups when diluted with
POPC~\cite{browning80,buldt81,roux90,roux91,scherer87}.
Unfortunately, none of the tested models correctly reproduce the modulation of POPS headgroup order
parameters with increasing amount of POPC in POPC:POPS mixtures (Fig. \ref{HGorderparametersPSvsPC}).
and we conclude that more accurate force fields are needed
to correctly describe the PC-PS headgroup interactions in MD simulations.
%Having this conclusion in every paragraph gets a bit repetive
